Design and Implementation of Probabilistic Programming Language Anglican

### 2017-02-21 paper/paper.tex
update

Anglican programs define probability distributions 
over sequences of annotated values, implicitly by means
of program execution.
Imagine an interpreter that runs the deterministic parts of a program 
according to the standard semantics, executes the \texttt{sample} form by
generating a random sample, and treats the $\texttt{observe}$ form by skip.
In addition to this execution, the interpreter also keeps a log that records
information about all the \texttt{sample} and \texttt{observe} forms encountered 
during execution. The information recorded for \texttt{sample} is 
a triple $(F,x,\alpha)$ of (i) a primitive probability distribution $F$, 
such as the standard normal, for which we have 
the probability density $p_F$; (ii) a value $x$ sampled from the distribution $F$; 
and (iii) an address $\alpha$ that uniquely and systematically identifies the random 
choice made. The information recorded for \texttt{observe} is a pair $(G,y)$ where $G$
is a primitive probability distribution as $F$ from above and $y$ is an 
observed value. Thus, a log is a sequence of triples $(F,x,\alpha)$ and 
pairs $(G,y)$.

A \textit{trace} $\pmb{x}$ is a sequence of triples $(F,x,\alpha)$. It is feasible if
it is precisely the triple part of the log of some execution. 
Such a feasible trace records all the random choices made during execution,
and uniquely determines the rest of the execution and its log, because all the non-sample
forms of Anglican programs are deterministic. In particular, each of these feasible traces
decides a sequence of pairs $(G_j, y_j)$ for observed values.
We call the sequence \textit{image of $\pmb{x}$} and denote it by~$\pmb{y}$.

A (almost surely terminating) probabilistic program defines a probability distribution 
over feasible traces $\pmb{x}$, which has the probability density 
$\pi(\pmb{x}) := \gamma(\pmb{x}) / Z$ where 
\begin{align}
  \gamma(\pmb{x}) 
  &:= 
  \prod_{i=1}^{\left|\pmb{x}\right|}
  p_{F_i}(x_i) \prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j}).
  \label{eqn:p-trace}
\end{align}
for the image $\pmb{y}$ of $\pmb{x}$, and $Z$ is the normalization constant $Z := \int \gamma(\pmb{x})d(\pmb{x})$
of $\gamma$.\footnote{The integral and the density use the default measure obtained by the standard extension of
a $\sigma$-finite measure on finite traces where the $\sigma$-finite measure itself is defined by
a countable sum of Lebesque and counting measures.}

>>>>

Anglican programs define probability distributions 
over sequences of values, implicitly by means of program execution.
A good way to understand this is to imagine the following 
interpreter of Anglican programs. Starting from a fixed initial state,
the interpreter runs the deterministic parts of a program 
according to the standard semantics, executes the \texttt{sample} form by
generating a random sample, and treats the $\texttt{observe}$ form by skip.
More importantly, the interpreter keeps a log that records
information about all the \texttt{sample} and \texttt{observe} forms encountered 
during execution. The information recorded for \texttt{sample} is 
a triple $(F,x,\alpha)$ of (i) a primitive probability distribution $F$, 
such as the standard normal, for which we have 
the probability density $p_F$; (ii) a value $x$ sampled from the distribution $F$; 
and (iii) an address $\alpha$ that uniquely and systematically identifies the random 
choice made. The information recorded for \texttt{observe} is a pair $(G,y)$ where $G$
is a primitive probability distribution as $F$ from above and $y$ is an 
observed value. Thus, a log is a sequence of triples $(F,x,\alpha)$ and 
pairs $(G,y)$.

One important property of logs is that they are determined by 
their projections to triples: when two logs project to the same
sequence of triples, they must be the same. This is because
the triples in a log contain all the information about random choices
made during execution and all the non-sample forms in Anglican programs
are deterministic. We define a \textit{trace} $\pmb{x}$ to be a sequence
of triples $(F,x,\alpha)$, and say that $\pmb{x}$ is feasible if
the trace is precisely the triple part of the log of some execution. 
Such a feasible trace uniquely determines the rest of the execution and its log.
In particular, it decides the pair part of the log, namely, 
a sequence of $(G_j, y_j)$ for observed values. We call this sequence \textit{image of $\pmb{x}$} 
and denote it by~$\pmb{y}$.

A (almost-surely terminating) probabilistic program defines a probability distribution 
over finite feasible traces $\pmb{x}$. The distribution has the probability density 
$\pi(\pmb{x}) := \gamma(\pmb{x}) / Z$ where 
\begin{align}
  \gamma(\pmb{x}) 
  &:= 
  \prod_{i=1}^{\left|\pmb{x}\right|}
  p_{F_i}(x_i) \prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j})
  \quad \mbox{for the image $\pmb{y}$ of $\pmb{x}$}, 
  \label{eqn:p-trace}
\end{align}
and $Z$ is the normalization constant $Z := \int \gamma(\pmb{x})d(\pmb{x})$
of $\gamma$.\footnote{The integral and the density use the default measure obtained by the standard extension of
a $\sigma$-finite measure on finite traces where the $\sigma$-finite measure itself is defined by
a countable sum of Lebesque and counting measures.}
